Research Notes

### 2018-04-22 writups/notes.tex
update writeup with recent musings

\subsection{Supressing Reflection}

@@

@@

\section{Incompressible, Stratified w/ Gravity, Nonlinear}

We consider the problem where $\rho_1 \sim \rho_0$, where we cannot make such a
perturbative expansion. We write down thus nonlinear fluid equations
\begin{equation}
    \begin{split}
        \pd{\rho}{t} + \vec{u} \cdot \p*{\vec{\nabla}\rho} &= 0,\\
        \vec{\nabla} \cdot \vec{u} &= 0,\\
        \pd{\vec{u}}{t} &= -\p*{\vec{u} \cdot \vec{\nabla}} \vec{u}
            - \frac{\vec{\nabla}P}{\rho} - g\hat{z}.
    \end{split}\label{eq:nonlin_incomp}
\end{equation}
Note that we cannot even subtract off hydrostatic equilibrium anymore since
$\rho$ can deviate greatly from the normal $\rho_0e^{-z/H}$!

>>>>

\section{Supressing Reflection}

@@

\subsection{First Order}

@@

\section{The Difficulty of the Incompressible, Nonlinear Problem}

We consider the problem where both advective terms are kept and $\rho_1 \sim
\rho_0$. We write down thus nonlinear fluid equations
\begin{equation}
    \begin{split}
        \pd{\rho}{t} + \vec{u} \cdot \p*{\vec{\nabla}\rho} &= 0,\\
        \vec{\nabla} \cdot \vec{u} &= 0,\\
        \pd{\vec{u}}{t} &= -\p*{\vec{u} \cdot \vec{\nabla}} \vec{u}
            - \frac{\vec{\nabla}P}{\rho} - g\hat{z}.
    \end{split}\label{eq:nonlin_incomp}
\end{equation}
Note that we cannot even subtract off hydrostatic equilibrium anymore since
$\rho$ can deviate greatly from the normal $\rho_0e^{-z/H}$!

\subsection{Rearrangement for Dedalus}

In order to simulate this at all in a spectral program, recall the way that
nonlinear PDEs are decomposed in a spectral code. Given a phase space $Q$,
consider first a system of PDEs expressible in form
\begin{equation}
    \dot{Q} + \bm{L}Q + f(Q) = g
\end{equation}
where $\dot{Q}$ is the time derivative, $\bm{L}$ is some linear operator, $f(Q)$
are nonlinear terms and $g$ is an arbitrary driving function. Spectral codes
work by re-casting the operators to be purely algebraic in an appropriate
spectral basis consisting of some $\varphi_i$ trial and $\xi_i$ test functions.
The \emph{tau spectral method} that Dedalus uses considers using the trial
functions as the test functions, so the effective decomposition that is
performed is
\begin{equation}
    \bra*{\varphi_i}\dot{Q} + \bm{L}Q \ket*{\varphi_j}
        = \dot{Q}_i + L_{ij}Q_q = \bra*{\varphi_i}-f(Q)  + g \ket*{\varphi_i}.
\end{equation}
It is then clear that $-f(Q) + g$ can just be treated as inhomogeneous source
terms, and over a short $\Delta t$ it can be approximated via $-f(Q_0) + g$.
This first-order inhomegeneous ODE then admits closed form exact solutions
$e^{-L_{ij}t}\p*{\dots}$.

One can imagine, for this matter, that $\rd{}{t}$ can be wrapped in the $L_{ij}$
operator and the timestepping computed implicitly; I believe this is what
Dedalus does, implicitly timestepping the left hand side and explicitly the
inhomogeneous terms.

Both of these procedures require $\bm{L}$ linear operator or $\rd{}{t} + \bm{L}$
to be full rank, or invertible. This is enforced analytically by requiring the
same number of BCs as derivatives, and numerically this is little different.
This is why it is possible to have components of $Q$ that are not differentiated
$\partial_t$ but still have well-defined time evolutions; the only condition
required is that the linear terms of the PDE make up a full-rank operator.
Hence, in our linear incompressible equations, we permitted use of $\vec{\nabla}
\cdot \vec{u}_1 = 0$ in lieu of a $\pd{P}{t}$ explicit equation of motion; so
long as the gauge for $P$ (which only appears in derivatives) is properly set,
the operator is full rank and numerically well-defined.

In the case of the full nonlinear equations though, it bears noting that $P$
only appears in nonlinear terms! Thus, the linear terms clearly cannot have full
rank since they fail to reference $P$ at all. We remedy this by re-casting the
momentum equation as
\begin{equation}
    \pd{\vec{u}}{t} + \frac{\vec{\nabla}P}{\rho_0}
        = -\vec{\nabla}P\p*{\frac{1}{\rho} - \frac{1}{\rho_0}}
            + \vec{g}.
\end{equation}
This ensures that, along with suitable gauge choice for $P$, the linear
operators on the left hand side of the equation have full rank. The remaining
equations
\begin{subequations}
    \begin{align}
        \pd{\rho}{t} &= -\vec{u} \cdot \vec{\nabla}\rho,\\
        \vec{\nabla} \cdot \vec{u} &= 0,
    \end{align}
\end{subequations}
remain unchanged.

\subsection{Pathology with $u_z$ BCs}

Our problem turns out to be ill-conditioned, the pathology arising from our
driving term in conjunction with the incompressibility constraint. Since we want
to model the propagation of waves from where they are excited, rather than
exciting them directly in our problem, we must drive our region of interest via
boundary conditions rather than forcing terms.

By driving via boundary conditions, we can only specify boundary conditions on
the variables being differentiated $\partial_z$ in our equations of motion,
since we enforce fully periodic BCs in $x$ via using a Fourier spectral
decomposition. Thus, we are restricted to enforcing BCs on either $\vec{u}$ or
$P$.

It turns out that by using BCs on $\vec{u}$, we run into a problem: the BCs we
specify on $u_z$ or its derivatives will in general affect grid points
$\partial_z u_{z, 1x}$ while those specified on $u_x$ can only affect
$\partial_x u_{x, 0x}$. This means that $\partial_x u_x$ will pick up
sharp jerks by being coupled to the BC, producing high frequency components in
$u_x$ as can be seen in \autoref{fig:agree_plots}. These come from enforcing the
divergence-free condition too literally at the boundary. Not that the figure was
produced with only a Dirichlet BC on $u_z$, no BCs on $u_x$, in a linear system
with absorbing BCs.
\begin{figure}[!h]
    \centering
    \includegraphics[width=0.7\textwidth]{../sims/2d_1_strat/agree_plots/sponge_2.png}
    \caption{Note the jaggedness in $u_x$ induced by a Dirichlet BC on
    $u_z$.}\label{fig:agree_plots}
\end{figure}

This problem seems to be remedied somewhat by instead enforcing Neumann BCs on
$u_z$, or even when allowing dissipation enforcing Neumann BCs on $\partial_z
u_z$ (the extra derivatives granted by $\nu \nabla^2 \vec{u}$ let us specify
higher order BCs). However, in the fully nonlinear system we still observe
divergences near $z=0$ regardless of how smooth of BCs we use on $u_z$.
Moreover, since the discontinuities come from the nature of the BC and not its
strength, we run into these instabilities regardless of how gradually we
introduce the BC (e.g.\ we could consider multiplying by $1 - e^{-t/\tau}$ to
gradually introduce the BC, but this turns out also to be unstable).

There are thus three possible solutions to this:
\begin{itemize}
    \item Consider BCs only on $P$, the only other choice of dynamical variable.

    \item Consider using a $\nabla^2 P$ equation instead of a $\vec{\nabla}
        \cdot \vec{u} = 0$ constraint. The correct choice of BCs in this case is
        discussed in Nordstrom et al, 2006.

    \item Change the system of equations to no longer consider an incompressible
        system; the anelastic system is an attractive alternative. This boasts
        the advantage of not having an algebraic constraint equation.
\end{itemize}

\section{Limits of the Incompressibility and Anelasticity Assumptions}

To get a better handle on the regimes of validity of the incompressible and
anelastic approximations, we return to the full fluid equations
\begin{equation}
    \begin{split}
        \pd{\rho}{t} + \vec{\nabla} \cdot \p*{\rho \vec{u}} &= 0,\\
        \pd{\vec{u}}{t} + \p*{\vec{u} \cdot \vec{\nabla}}\vec{u} +
            \frac{\vec{\nabla}P}{\rho} - \vec{g} &= 0.
    \end{split}
\end{equation}
Furthermore, we assume small perturbations $\delta P = c_s^2 \delta \rho$ where
$c_s^2$ is the sound speed, also the velocity at which an adiabatic perturbation
is restored. We drop the $-\vec{g}$ from the analysis assuming that it is
largely negated by an appropriate background $P$.

Given these considerations, we consider in what limit we may presume the fluid
to be incompressible. Consider if $\abs*{\vec{u}} \ll c_s$, then for some
characteristic frequency $\omega$ and wavenumber $k$ the momentum equation
scales as $\omega u + ku^2 + c_s^2 \delta \rho/\rho = 0$. Comparing scales it is
thus clear that $\delta \rho \ll \rho$ in this limit. Then, considering the
mass continuity equation, we recognize that the scalings are $\omega \delta \rho
+ \rho \vec{\nabla} \cdot \vec{u} + k\delta \rho u = 0$. Thus, $\vec{\nabla}
\cdot \vec{u} = 0$ necessarily, since $\rho$ is much larger than the other two
terms, and we obtain the incompressible mass continuity equation from the other
two terms. While this is obviously not a very careful treatment, it should
suffice to convince us that $u \ll c, \delta \rho \ll \rho, \vec{\nabla} \cdot
\vec{u} = 0$ are all roughly equivalent conditions.

On the other hand, when may we drop the total derivative terms? This seems to be
a slightly different criterion, and indeed we may compare $\pd{\vec{u}}{t} +
\p*{\vec{u} \cdot \vec{\nabla}} \vec{u} \sim \omega \vec{u} + ku^2$. Thus, the
second term is dominated by the first term when $\vec{u} \ll \vec{v}_{ph}$ where
$\vec{v}_{ph} \equiv \frac{\omega}{k}\hat{k}$ is the phase velocity. Recall that
for our stratified waves that $\omega \leq N = \sqrt{g/H}$ as a point of
reference. So we require $u \ll v_{ph}$ to be able to drop the total derivative
terms.

Since in general $v_{ph} \ll c_s$, we thus identify that there are three
regimes. The first, $u \ll v_{ph}$, we solve the completely linear problem.
Then, for $v_{ph} \lesssim u \ll c_s$, we must begin to consider the advective
terms but are still allowed to assume $\rho_1 \ll \rho_0$, an assumption that is
only violated when $u \sim c_s$.

If we look ahead to the anelastic system derivation, we can identify that the
anelastic approximation corresponds to assuming $c_s^2 \gg uv_{ph}$, making no
assertion of the scale between $u, v_{ph}$. Thus, we can also consider the
anelastic approximation with advective terms just fine in teh regime $v_{ph}
\lesssim u$, and if this is sufficient to induce wave breaking then we may never
need to consider the $\rho_1 \sim \rho_0$ enormous perturbations.
